Equation-free analysis of a dynamically evolving multigraph

### 2016-02-10 report/multigraph_manuscript.tex
Moor manuscript work

PCA is used to linearly project data into subspaces that capture the
directions along which the data varies most
\cite{jolliffe_principal_2014}. Given some matrix
$X \in \mathbb{R}^{n \times m}$ in which each of the $m$ column
vectors represent a different collection of the $n$ system variables,
PCA computes a reduced set of $k$ ``principal components''
$z_i \in \mathbb{R}^n$ with $k < n$ that constitute an optimal basis
for the data in the sense that, among linear, $k$-dimensional embeddings, $W = Z^TX$
captures the maximum possible variance in the dataset. \\

This method has found wide application, but can suffer from its
inability to uncover nonlinear relationships among the input
variables. Indeed, many datasets will not lie along some hyperplane
embedded in $\mathbb{R}^n$, but rather will form a nonlinear manifold
throughout the space. \\

Theoretical results in \cite{rath_time_2012} state that, for a given network
size $n$, the final stationary state depends only on the number of
edges present $m$ and the model parameter $\kappa$. To assess both the
validity of our graph embedding technique and the usefulness of the
PCA in the context of complex networks, we generated a dataset of
stationary states over a range of $m$ and $\kappa$ values by directly
running the model over $O(n^3)$ steps ($N$ values of each parameter
were taken, for a total of $N^2$ networks). Each resulting
graph $G(m_i, \kappa_j) = G_{ij}$ was then embedded into
$\mathbb{R}^8$ by counting the number of times each of the subgraphs
shown in Fig. (\ref{fig:motifs}) appeared in the network. Thus
$\gamma(G_{ij}) = v_{ij} \in \mathbb{R}^8$. We then proceeded to
perform a principal component analysis on this collection of vectors
$\{v_{ij}\}_{i,j=1}^N$. Surprisingly, this linear technique succeeded in
uncovering a two-dimensional embedding of the dataset corresponding to
the two underlying parameters $m$ and $\kappa$, as shown in
Fig. (\ref{fig:pca}). This suggests that, given some final network
state $G(m, \kappa)$, by projecting its embedding onto these first two
principal components one could obtain a reasonable approximation of
the hidden parameter $\kappa$.

@@

Additionally, we were interested in embeddings of different
model trajectories. This dataset was generated by sampling two
different model simulations at even intervals as they evolved ($N$
points were sampled from each trajectory). The parameter values $m$
and $\kappa$ were held constant, but one graph was initialized as an
Erd\H{o}s-R\'{e}nyi random graph (Fig. (\ref{fig:erdos-init})), while
the other was initialized as a lopsided graph
(Fig. (\ref{fig:lopsided-init})). Letting $G_e(t)$ refer to the
Erd\H{o}s-R\'{e}nyi-initialized system at step $t$, and $G_l(t)$ the
lopsided-initialized system, the embedding $\gamma$ was used to create
points $\gamma(G_e(t)) = v_e(t) \in \mathbb{R}^8$ and similarly
$\gamma(G_l(t)) = v_l(t) \in \mathbb{R}^8$. \\

@@

\begin{figure}[h!]
  \vspace{-5mm}
  \centering
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{dmaps-3d-convergence-a}
    \subcaption{\label{fig:dmaps-results-regular}}
  \end{subfigure} %
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{dmaps-3d-convergence-zoom-a}
    \subcaption{\label{fig:dmaps-results-zoom}}
  \end{subfigure}%
  \caption{DMAP of motif-based embeddings from two separate simulation
    trajectories: (a) embedding of two trajectories on different
    timescales. While the trajectories are initially embedded in
    separate regions due to their different initial conditions,
    they're eventually mapped close together as they evolve to the
    same coarse stationary state. Final states are circled, and point
    size grows as more steps are taken in each trajectory. (b)
    Enlarged view of the final points in the previous plot, revealing
    the approach to a similar final state. Here, the average of the
    final fifty points is circled. Due to the stochastic nature of
    the system, the final embeddings randomly fluctuate about the
    shared stationary state. \label{fig:dmaps-results}}
\end{figure}

\begin{figure}[h!]
  \vspace{-5mm}
  \centering
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{dmaps-2d-rho-a}
    \subcaption{\label{fig:dmaps-rho}}
  \end{subfigure} %
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{dmaps-2d-kappa-a}
    \subcaption{\label{fig:dmaps-kappa}}
  \end{subfigure}%
  \caption{DMAP of motif-based embeddings from collection of different
    simulations run with different parameter values: (a) coloring the
    two-dimensional DMAPS embedding with $\rho$ and (b) coloring the
    two-dimensional DMAPS embedding with $\kappa$. As with PCA, DMAPS
    uncovered the variables governing the stationary state, known from
    Bal\'{a}zs' paper \label{fig:dmaps-rk}}
\end{figure}

>>>>

PCA is used to linearly project data into subspaces that capture the
directions along which the data varies most
\cite{jolliffe_principal_2014}. Given some matrix
$X \in \mathbb{R}^{n \times m}$ in which each of the $m$ column
vectors $x_i$ represent a different collection of the $n$ system
variables (i.e. a different data point), PCA computes a reduced set of
$k<n$, orthogonal ``principal components'' $z_i \in \mathbb{R}^n$ that
constitute an optimal basis for the data in the sense that, among
linear, $k$-dimensional embeddings, $w_i = Z^Tx_i$ captures the
maximum possible variance in the dataset, where
$Z = \begin{bmatrix} | & | & & | \\ z_1 & z_2 & \hdots & z_k \\ | & | & &
  | \end{bmatrix}$. This method has found wide application, but can suffer from its
inability to uncover nonlinear relationships among the input
variables. Indeed, many datasets will not lie along some hyperplane
embedded in $\mathbb{R}^n$, but rather will form a nonlinear manifold
throughout the space. \\

Theoretical results in \cite{rath_time_2012} state that, for a given
network size $n$, the final stationary state depends only on the
number of edges present $m$ and the model parameter $\kappa$. To
assess both the validity of our graph embedding technique and the
usefulness of the PCA in the context of complex networks, we generated
a dataset of stationary states over a range of $m \in [50, 5,000]$ and
$\log(\kappa) \in [0, 2]$ values by directly running the model over
$2n^3$ steps ($N=30$ values of each parameter were taken, for a total
of $900$ networks). We fixed the network size to $n=50$ vertices. Each
resulting graph $G(m_i, \kappa_j) = G_{ij}$ was then embedded into
$\mathbb{R}^8$ by counting the number of times each of the subgraphs
shown in Fig. (\ref{fig:motifs}) appeared in the network. Thus
$\gamma(G_{ij}) = v_{ij} \in \mathbb{R}^8$. We then proceeded to
perform a principal component analysis on this collection of vectors
$\{v_{ij}\}_{i,j=1}^N$. Surprisingly, the first two principal
components $z_1$ and $z_2$ succeeded in uncovering a two-dimensional
embedding of the dataset corresponding to the two underlying
parameters $m$ and $\kappa$, as shown in Fig. (\ref{fig:pca}) in which
the data is projected onto these to vectors. This suggests that, given
some final network state $G(m, \kappa)$, by projecting its embedding
onto these first two principal components one could obtain a
reasonable approximation of the hidden parameter $\kappa$.

@@

Additionally, we were interested in embeddings of different model
trajectories. This dataset was generated by sampling two different
model simulations at even intervals as they evolved ($N=2000$ points were
sampled from each trajectory). The parameters $n$, $m$ and $\kappa$
were held constant at $200$, $20100$ and $1.0$ respectively, but one
graph was initialized as an Erd\H{o}s-R\'{e}nyi random graph
(Fig. (\ref{fig:erdos-init})), while the other was initialized as a
lopsided graph (Fig. (\ref{fig:lopsided-init})). Every $500$ steps the
graph would be recorded till $N$ snapshots were taken of each
trajectory, for a total of $1000000$ steps. Letting $G_e(t)$ refer
to the Erd\H{o}s-R\'{e}nyi-initialized system at step $t$, and
$G_l(t)$ the lopsided-initialized system, the embedding $\gamma$ was
used to create points $\gamma(G_e(t)) = v_e(t) \in \mathbb{R}^8$ and
similarly $\gamma(G_l(t)) = v_l(t) \in \mathbb{R}^8$. \\

@@

\begin{figure}[h!]
  \vspace{-5mm}
  \centering
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{dmaps-3d-convergence-a}
    \subcaption{\label{fig:dmaps-results-regular}}
  \end{subfigure} %
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{dmaps-3d-convergence-zoom-a}
    \subcaption{\label{fig:dmaps-results-zoom}}
  \end{subfigure}%
  \caption{DMAP of motif-based embeddings from two separate simulation
    trajectories: (a) embedding of two trajectories starting from
    different initial conditions. While the trajectories are initially
    mapped to separate regions, their embeddings eventually overlap as
    they evolve to the same coarse stationary state. Final states are
    circled, and point size grows as more steps are taken in each
    trajectory. (b) Enlarged view of the final points in the previous
    plot, revealing the approach to a similar final state. Here, the
    average of the final fifty points is circled. Due to the
    stochastic nature of the system, the final embeddings randomly
    fluctuate about the shared stationary
    state. \label{fig:dmaps-results}}
\end{figure}

\begin{figure}[h!]
  \vspace{-5mm}
  \centering
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{dmaps-2d-rho-a}
    \subcaption{\label{fig:dmaps-rho}}
  \end{subfigure} %
  \begin{subfigure}{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{dmaps-2d-kappa-a}
    \subcaption{\label{fig:dmaps-kappa}}
  \end{subfigure}%
  \caption{DMAP of motif-based embeddings from collection of different
    simulations run with different parameter values: (a) coloring the
    two-dimensional DMAPS embedding with $\rho$ and (b) coloring the
    two-dimensional DMAPS embedding with $\kappa$. As with PCA, DMAPS
    uncovered the variables governing the stationary state, known from
    \cite{rath_time_2012} \label{fig:dmaps-rk}}
\end{figure}
